\texttt{Supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-03-20 doc/incertae_sedis_latex2.tex
Update ms.

@@

The phrase \textquotedblleft \emph{incertae sedis}\textquotedblright ,
which is a Latin phrase meaning ``uncertain seat'', indicates that
a taxon is of uncertain placement. Such groups arise in practice when
they are classified within a parent group, such as Fungi, but the
precise location within the parent group is uncertain. Such a taxon
would then be attached as a child of the parent group, but labeled
as \emph{incertae sedis}''.

The OpenTree project seeks to build a comprehensive supertree covering
all of life. The approach is to combine information in published phylogenies
with a comprehensive taxonomy that supplies taxon names \citep{redelings2017supertree}.
We describe an extension to this supertree method so that we can resolve
taxonomic uncertainty by using published phylogenies to place \emph{incertae
sedis} taxa. We describe an updated semantics for conflict between
phylogeny edges and taxonomy edges and an updated semantics for assigning
taxonomy names to nodes in the synthetic tree.

The ability to place taxa within the taxonomy without breaking containing
taxa enables us to include thousands of new taxa that were previously
filtered out to avoid broken taxa. It also enables us to use published
phylogenies to update the taxonomy by extending taxa to include \emph{incertae
sedis} taxa placed within them. This makes substantial progress towards
our goal of complete representation of taxa. It also enables us to
include extinct taxa, since many of these taxa are incertae sedis.

@@

We briefly describe the supertree algorithm proposed in \citet{redelings2017supertree}.
This supertree algorithm takes as input a ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
of rooted input trees and a single rooted taxonomy tree $\taxonomy$.
The taxonomy tree is ranked below any of the input trees. Each node
$i$ in the taxonomy tree is associated a taxon name $n(i)$ in the
set of taxon names $\mathcal{N}$, and each taxon name is corresponds
to only one node of the taxonomy tree. Thus the taxonomy is comprehensive
and contains all taxa. In contrast, input phylogenies only have labels
on their leaf nodes, and generally include a small fraction of the
total number of taxa. Furthermore, taxa in input trees may correspond
to higher taxa on the taxonomy tree, and multiple tips may be labeled
with the same taxon name. \citet{redelings2017supertree} described
a method called 'exemplification' that reduces this problem to a simpler
form where all taxon names of input trees (i) correspond to leaf taxa
and (ii) occur only once in an input tree. For the purposes of this
paper, we therefore assume that this is the case.

@@

Supertrees for each subproblem are obtained by greedily constructing
a set of rooted splits $\mathcal{C}$ that are jointly consistent
and correspond to branches of the input trees. We construct an ordered
list of splits $S(T_{i})$ by walking the branches of tree $T_{i}$
in post-order and appending the corresponding splits to the list.
These lists combined to form the list $\Sigma=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$,
where '$+$' denotes concatenation. The set $C$ is initialized to
the empty set, and we then iteratively consider each split in the
list $\Sigma$ and insert it into $C$ if the resulting set remains
jointly compatible. The BUILD algorithm is used both to check compatibility,
and to construct a compatible tree from the final set $C$.

@@

In order to incorporate \emph{incertae sedis }taxa into a supertree
analyses, we first tackle the issue of the semantics of \emph{incertae
sedis} taxa. We discuss various different things the term can mean
within the literature, and settle on a specific usage here. We discuss
how to move from an operational ``you can move this guy here'' semantics
to a splits-based semantics. This affects both what it means for a
taxonomy edge to conflict with the synthesis tree, and what it means
to place a taxon name on a particular node in the synthesis tree.
Finally, after a supertree has been constructed, we consider how to
interpret the supertree as (perhaps) placing an \emph{incertae sedis}
taxon $A$ within its sister taxon $B$.

\subsection{Notation}

We extend this framework by adding a set $\mathcal{I}\subseteq\mathcal{N}$
of labels that have the \emph{incertae sedis }property. The taxonomy
consists of the labels $\mathcal{N}$, the taxonomy tree $\taxonomy$,
and the \emph{incertae sedis} property $\mathcal{I}$ of taxon names.

\subsection{Usage in the literature}

@@

\subsubsection{Splits}

Each edge of a standard tree divides the tip taxa $\mathcal{L}$ into
two groups: the include set $\mathcal{I}(e)$ which does not contain
the root, and the exclude set $\mathcal{E}(e)$ which does contain
the root. Such a split may be written
\[
\mathcal{I}(e)|\bullet\mathcal{E}(e).
\]
where the $\bullet$ indicates the root. If no taxa are \emph{incertae
sedis}, then the exclude set for a node is just the total tip set
minus the include set for the node:
\begin{align*}
\mathcal{E}(n) & =\mathcal{L}-\mathcal{I}(n).
\end{align*}

For a node $n$ on the tipward side of an edge $e$, we may also write
$\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for
$\mathcal{E}(e)$. We consider the exclude set of the root node to
be empty, and the include set of the root node to contain all tip
taxa:
\begin{align*}
\mathcal{E}(root) & =\{\}\\
\mathcal{I}(root) & =\mathcal{L}.
\end{align*}
In this case the root node stands for an edge that connects the root
to the root's parent.

@@

An \emph{incertae sedis} taxon can be moved into any of the descendants
of its siblings. Therefore, the exclude set $\mathcal{E}(n)$ for
node $n$ should not include any of its siblings marked \emph{incertae
}sedis\emph{,} but should exclude the descendants of its other siblings\emph{.}
Additionally, the exclude $\mathcal{E}(n)$ should contain the children
of all the ancestors of $n$, unless those children are incertae sedis.

We may also consider what happens if an \emph{incertae sedis} taxon
$A$ is placed within a sister group. Suppose $A$ is placed next
to a sibling $B$. If $B$ is \emph{incertae sedis}, then it should
be possible to place $B$ within $A$. As a result, $A$ should not
exclude descendants of $B$ than can only be reached from $A$ by
passing through a node marked \emph{incertae sedis. }

@@

However, in the \emph{incertae sedis }framework we must raise the
question of whether one name could apply to multiple nodes, or whether
multiple names could apply to one node.

\textbf{BDR:} \emph{It is well known that monotypic taxa are indistinguishable
if you consider only splits on leaf labels, but are distinguishable
if you consider splits on all node labels. It seems that some (all?)
of the problems with assigning multiple names to the same nodes actually
comes from the fact that moving IS taxa can leave the parent as a
degree-2 node. The fact that propinquity handles monotypic taxa indicates
that we actually implicitly consider all taxa to have labels. It is
possible that a trivial extension to the sub-problem solved could
thus handle monotypic taxa and issues with mapping 2 names to one
node with incertae sedis taxa.}

\subsubsection{Multiple nodes that fit one name}

Suppose the taxonomy is (((A1,A2)A,B)AB,C,D{*}) and the solution tree
is (((A1,A2)x,D{*})y,B)AB,C). In this case the name $A$ leads to
the split S(A) = A1 A2 | B C root. This name can apply to both the
node x and the node y.

\textbf{Solution:} In this case, we find the most tipward node and
attach the name to this node.

@@

>>>>

\section{New Introduction}

Supertree methods combine a collection of input trees with different
taxon sets into a single tree on the combined taxon set. While such
methods are usually applied to leaf-labeled phylogenies that have
been estimated from DNA sequence data, we additionally seek to include
taxonomic information. We thus require that one of the input trees
is the taxonomy tree. The taxonomy has a number of features not found
in the input trees, such as having labels for internal nodes, containing
monotypic nodes, and containing \emph{incertae sedis} nodes. We refer
to labels of taxonomy nodes as ``taxa'', and require that labels
and taxonomy nodes have a one-to-one correspondence. We additionally
require that the taxonomy be comprehensive and contain all labels.

One of the problems that arises when attempting to include taxonomic
information is the problem of \emph{incertae sedis }taxa. \emph{Incertae
sedis} taxa are taxa that have unknown taxonomic placement. The common
interpretation of such taxa is that they may not be place further
root-ward, but may be placed within their sibling taxa. This frequently
occurs when a specimen is identified to a taxonomic level such as
the family level but not further. It also occurs when it is unclear
whether a genus forms its own family, or is part of an existing family.
(Mark rewrite?)

@@

The phrase \textquotedblleft \emph{incertae sedis}\textquotedblright ,
which is a Latin phrase meaning ``uncertain seat'', indicates that
a taxon is of uncertain placement. Such groups arise in practice when
they are classified within a parent group, such as Fungi, but the
precise location within the parent group is uncertain. Such a taxon
would then be attached as a child of the parent group, but labeled
as \emph{incertae sedis}.

The Open Tree of Life project seeks to build a comprehensive supertree
covering all of life. The approach is to combine information in published
phylogenies with a comprehensive taxonomy that supplies taxon names
\citep{redelings2017supertree}. We describe an extension to this
supertree method so that we can resolve taxonomic uncertainty by using
published phylogenies to place \emph{incertae sedis} taxa. We describe
an updated semantics for conflict between phylogeny edges and taxonomy
edges and an updated semantics for assigning taxonomy names to nodes
in the synthetic tree.

Our approach to \emph{incertae sedis} taxa enables us to use published
phylogenies to update the taxonomy by extending taxonomic groups to
include \emph{incertae sedis} taxa placed within them. The ability
to update the taxonomy is important, since the placement of \emph{incertae
sedis} taxa creates artificial conflicts with a non-updated taxonomy.
Thus, the ability to correctly assess compatibility with the taxonomy
when the taxonomy contains \emph{incertae sedis taxa} allows us to
include thousands of new taxa that were previously filtered out to
avoid conflict. It also enables us to include extinct taxa, since
many of these taxa are \emph{incertae sedis}. This makes substantial
progress towards our goal of completely representation of taxa.

@@

We briefly describe the supertree algorithm proposed in \citet{redelings2017supertree}.
This supertree algorithm takes as input a ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
of rooted input phylogenies and a single rooted taxonomy tree $\taxonomy$.
Each node in the taxonomy tree is associated with a taxon name in
the set of taxon names $\mathcal{N}$, and each taxon name is corresponds
to only one node of the taxonomy tree. We use \emph{$\mathcal{L}$
}to indicate taxon names corresponding to leaf nodes of the taxonomy
tree. Correspondence between the input trees and the taxonomy is established
by annotating leaf nodes of each input tree with taxon names. For
purposes of conflict resolution, the taxonomy tree is ranked below
all of the input trees.

\paragraph{Exemplification}

While the taxonomy is comprehensive and contains all taxa, input phylogenies
only have labels on their leaf nodes, and generally include a small
fraction of the total number of taxa. Furthermore, taxa in input trees
may correspond to higher taxa on the taxonomy tree, and multiple tips
may be labeled with the same taxon name. \citet{redelings2017supertree}
described a method called 'exemplification' that reduces this problem
to a simpler form where all taxon names of input trees (i) correspond
to leaf taxa and (ii) occur only once in an input tree. For the purposes
of this paper, we therefore assume that this is the case.

\paragraph{Problem decomposition}

@@

\paragraph{Splits-based synthesis method}

Supertrees for each subproblem are obtained by greedily constructing
a set of rooted splits $\mathcal{C}$ that are jointly consistent
and correspond to branches of the input trees. We construct an ordered
list of splits $S(T_{i})$ by walking the branches of each tree $T_{i}$
in post-order and appending the corresponding splits to the list.
These lists are combined to form the list $\Sigma=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$,
where '$+$' denotes concatenation. The set $C$ is initialized to
the empty set, and we then iteratively consider each split in the
list $\Sigma$ and insert it into $C$ if the resulting set remains
jointly compatible. The BUILD algorithm is used both to check compatibility,
and to construct a compatible tree from the final set $C$.

@@

We extend this framework by adding a set $\mathcal{U}\subseteq\mathcal{N}$
of labels that have the \emph{incertae sedis }property. The extended
taxonomy then consists of the taxonomy tree $\taxonomy$, the labels
$\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis }taxa.

@@

In order to apply supertree methods to rooted trees with \emph{incertae
sedis} taxa, we first propose a semantics of \emph{incertae sedis
}in terms of rooted splits.

We define rooted splits by noting that each edge of a tree divides
the tip taxa $\mathcal{L}$ into two groups: the include set $\mathcal{I}(e)$
which does not contain the root, and the exclude set $\mathcal{E}(e)$
which does contain the root. Such a split may be written
\[
\mathcal{I}(e)|\bullet\mathcal{E}(e).
\]
where the $\bullet$ indicates the root. If no taxa are \emph{incertae
sedis}, then the exclude set for a node is just the total tip set
minus the include set for the node:
\begin{align*}
\mathcal{E}(n) & =\mathcal{L}-\mathcal{I}(n).
\end{align*}

For a node $n$ on the tip-ward side of an edge $e$, we may also
write $\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$
for $\mathcal{E}(e)$.

@@

An \emph{incertae sedis} taxon can be moved into any of the descendants
of its siblings. Therefore, the exclude set $\mathcal{E}(n)$ for
node $n$ should not include any of its siblings marked \emph{incertae
}sedis\emph{,} but should exclude the descendants of its other siblings\emph{.}
Additionally, the exclude $\mathcal{E}(n)$ should contain the children
of all the ancestors of $n$, unless those children are \emph{incertae
sedis}.

We may also consider what happens if an \emph{incertae sedis} taxon
$A$ is placed within a sister group. Suppose $A$ is placed next
to a sibling $B$. If $B$ is also \emph{incertae sedis}, then it
should be possible to place $B$ within $A$. As a result, $A$ should
not exclude descendants of $B$ than can only be reached from $A$
by passing through a node marked \emph{incertae sedis. }

@@

However, in the \emph{incertae sedis} framework, it is possible for
one name to apply to multiple nodes. For example, in Figure \ref{fig:One-name-can},
the name $A$ can apply to multiple nodes. Here the name $A$ corresponds
to the split $A1\,A2\,|\bullet B\,C$, leaving out $D$ since it is
\emph{incertae sedis}. The two nodes $x$ and $y$ imply the splits
$A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet B\,C$ respectively,
and both of these splits imply the split $A1\,A2\,|\bullet B\,C$,
so the name $A$ can apply to both $x$ and \footnote{Define ``implies'', as in the propinquity paper: $A_{1}|\bullet A_{2}\implies B_{1}|\bullet B_{2}$
means $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.}. This cannot happen without\emph{ incertae sedis} taxa except at
monotypic nodes. When faced with a choice about where to place a name,
we find the most tip-ward node where the name can apply and attach
the name to this node. This corresponds to a conservative choice about
whether \emph{incertae sedis} taxa should be added to an existing
taxon.
\begin{figure}
\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/tax-shaded}

}\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/solution-shaded}

\caption{\label{fig:One-name-can}One name can be consistent with multiple
nodes.}
\end{figure}

It is also possible for multiple names to apply to a single node.

<--------------------- resume editting here ------------->

\emph{}

@@

\subsubsection{Usage in the literature}
